---
title: "20190626_AllBloodDarmanJunctions"
author: "Ben Fair"
date: "6/26/2019"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

This analysis was a meta-analysis of all the blood or bone marrow samples I could find in snaptron splice-junction count tables. This amounts to ~6000 samples across TCGA, SRA2, and GTEx. The count table I will load is only the subset of junctions within 1000bp of a sf3b1 differentially spliced intron from Darman et al.

```{r libraries}
library(data.table)
library(dplyr)
library(ggplot2)
library(reshape2)
library(readxl)
library(ggrepel)
library(tidyverse)
library(gplots)
library(knitr)
```

```{r}
gsub2 <- function(searchstring, replacestring, query){
  # function like gsub but will return NA if no match
  ind <- grep(searchstring,query,value = FALSE)
  query <- gsub(searchstring, replacestring, query, perl=T)
  query[-ind] <- NA
  query
}


Numers <- '../output/snakemake_out_sf3b1_blood_DarmanJuncs/CountTableNumerators.gz'
Denoms <- '../output/snakemake_out_sf3b1_blood_DarmanJuncs/CountTableDenominators.gz'

NumeratorCountTable <- fread(paste0("gunzip -c ", Numers), sep=" ", header=TRUE, data.table=FALSE)
DenominatorCountTable <- fread(paste0("gunzip -c ", Denoms), sep=" ", header=TRUE, data.table=FALSE)

PSI_Table <- NumeratorCountTable[-1]/DenominatorCountTable[-1]
row.names(PSI_Table) <- gsub(":clu_.+", "", NumeratorCountTable$chrom)
colnames(PSI_Table) <- gsub(".+\\.(.+$)", "\\1", colnames(PSI_Table), perl=T)

#Replace NaNs from 0/0 operation
PSI_Table[is.na(PSI_Table)] <- 0

# DifferentialJunctionsList <- read.table('/Users/benfair/Downloads/Seiler_S3B1InBRCA_DifferentialJunctions.txt', header=F)
DifferentialJunctionsList <- read.table('../data/other_snakemake_data/DarmanDiffJunctions.LiftedToHg38.bed', header=F)

# Make new id, for easier use in later filtering. Note that coordinates may need to add 1
DifferentialJunctionsList$junc_id <- paste(DifferentialJunctionsList$V1, DifferentialJunctionsList$V2 + 1, DifferentialJunctionsList$V3 + 1, sep=":")

# PCA using PSI
PCResults <- (PSI_Table) %>%
  filter(rownames(.) %in% DifferentialJunctionsList$junc_id) %>%
  t() %>%
  prcomp(center=T, scale=T) 


#Read in metadata for SRA2
SRA2Metadata <- fread('../output/snakemake_out_sf3b1_blood_DarmanJuncs/SnaptronMetadata.SRA2.txt.gz', header=TRUE, sep='\t', na.strings = "NA", stringsAsFactors = F)
SRA2Metadata$rail_id <- as.character(SRA2Metadata$rail_id)
SRA2Metadata$source <- "SRA2"

#Read in metadata for GTEX
GTEXMetadata <- read.delim('../output/snakemake_out_sf3b1_blood_DarmanJuncs/SnaptronMetadata.GTEX.txt.gz', header=TRUE, sep='\t', na.strings = "NA", stringsAsFactors = F)
GTEXMetadata$rail_id <- as.character(GTEXMetadata$rail_id)
GTEXMetadata$source <- "GTEX"

#Read in metadata for TCGA
TCGAMetadata <- read.delim('../output/snakemake_out_sf3b1_blood_DarmanJuncs/SnaptronMetadata.TCGA.txt.gz', row.names=1, na.strings = "NA", stringsAsFactors = T, header=T)

# truncated TCGA barcodes are easier to match and merge with Seiller et al metadata
TCGAMetadata$TruncatedTCGA <- gsub("-[0-9]{2}[A-Z]-[0-9]{2}[A-Z]", "", TCGAMetadata$gdc_cases.samples.portions.analytes.submitter_id)
TCGAMetadata$rail_id <- row.names(TCGAMetadata)
TCGAMetadata$source <- "TCGA"

# Read in Mutations list metadata from Seiller et al and add to metadata
MutationList <- data.frame(read_excel("../data/other_snakemake_data/SeillerEtAl_MutationList.xlsx"))
MutationList$TruncatedTCGA <- gsub("-[0-9]{2}[A-Z]-[0-9]{2}[A-Z]-\\w{4}-\\w{2}$", "", MutationList$Tumor_Sample_Barcode)

# Add labels to help distinguish SF3B1 hotspot mutations from other SF3B1 mutations in plot
MutationList$Mutant_aa <- paste(MutationList$Hugo_Symbol, MutationList$Protein_position, MutationList$Amino_acids, sep=":")
MutationList$Mutant_aa[which(MutationList$Hugo_Symbol!="SF3B1")] <- NA


Metadata <- MutationList %>%
  filter(Amino_acids != ".") %>%
  distinct(TruncatedTCGA, .keep_all=TRUE) %>%
  right_join(TCGAMetadata, by = "TruncatedTCGA", copy = FALSE) %>% 
  full_join(SRA2Metadata, by = c("junction_coverage", "junction_avg_coverage", "rail_id", "source")) %>%
  full_join(GTEXMetadata, by = c("junction_coverage", "junction_avg_coverage", "rail_id", "source")) %>%
  filter(rail_id!="1") %>% 
  mutate(sf3b1status = gsub2('.+?\\"(sf3b1.+?:.+?)\\".+', "\\1", characteristics.y)) %>%
  unite("Combined.sf3b1", sf3b1status, Mutant_aa)

Metadata$Combined.sf3b1[which(Metadata$Combined.sf3b1=="NA_NA")] <- NA

BloodContrast <- read.table("../data/other_snakemake_data/GTEX_Blood_sf3b1.groups", col.names = c("file", "groups"))
BloodContrast$railid <- gsub(".+\\.(.+$)", "\\1", BloodContrast$file, perl=T)

# Rename rows for easier merging later
PCResults.df <- as.data.frame(PCResults$x)
PCResults.df$rail_id <- sub("juncfiles\\..+?\\.", "", rownames(PCResults.df))
ToPlot <- left_join(PCResults.df, Metadata, by = "rail_id")

ggplot(ToPlot, aes(x=PC1, y=PC2, label=Combined.sf3b1, color=source)) + 
  geom_point(alpha=1) + 
  geom_text_repel(size=2, color="black") +
  theme(legend.title = element_blank(), legend.text=element_text(size=6)) +
  theme_bw()

```



Ok try to get matched wildtype and mutants from same studies and balanced for other factors

```{r}
StudiesWithMutants <- ToPlot %>%
  filter(PC1<=-10 & PC2<=-20 & !is.na(study_ID)) %>%
  select(study_ID) %>% distinct() %>% pull() %>% length()

ToPlot %>% filter(study_ID %in% StudiesWithMutants) %>%
ggplot(aes(x=PC1, y=PC2, label=Combined.sf3b1)) + 
  geom_point(alpha=1) + 
  geom_text_repel(size=2, color="black") +
  theme(legend.title = element_blank(), legend.text=element_text(size=6)) +
  theme_bw()

PoolToPickSamples <- ToPlot %>%
  mutate(CovariateInfo=paste(study_ID, cell_type, tissue,	cell_line)) %>%
  filter(junction_coverage>0)

Cases <- PoolToPickSamples %>%
  filter(PC1<=-10 & PC2<=-20 & !is.na(study_ID)) %>% 
  mutate(Status="Case") %>%
  sample_n(21) %>%
  select(rail_id, CovariateInfo, Status)

Controls <- PoolToPickSamples %>% filter(PC1>=10 & PC2>=-10) %>%
  filter(study_ID %in% StudiesWithMutants) %>%
  # sample_n(95) %>%
  mutate(Status="Control") %>%
  select(rail_id, CovariateInfo, Status)

#Check that study is balanced with regards to balancing covariates across cases and controls
rbind(Cases, Controls) %>%
  ggplot(aes(x=CovariateInfo, color=Status)) +
  geom_bar() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=3))
  
```

Write out samples for new differential splicing contrast with leafcutter (groups file, junctions file)

```{r}
rbind(Cases, Controls) %>%
  mutate(fullSampleName=paste0(".project2.yangili1.snaptron.processed_to_bedfiles.SRA2.", rail_id)) %>%
  select(fullSampleName, Status) %>%
  write.table("../data/other_snakemake_data/MostlyMarrowContrast.groups", sep='\t', quote=F, col.names = F, row.names = F)
  
rbind(Cases, Controls) %>%
  mutate(fullSampleName=paste0("/project2/yangili1/snaptron/processed_to_bedfiles/SRA2/", rail_id, ".junc")) %>%
  select(fullSampleName) %>%
  write.table("../data/other_snakemake_data/MostlyMarrowContrast.juncfiles", sep='\t', quote=F, col.names = F, row.names = F)
```


