---
title: "20190626_ds_sf3b1_blood"
author: "Ben Fair"
date: "6/25/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

```{r libraries}
library(data.table)
library(dplyr)
library(ggplot2)
library(reshape2)
library(readxl)
library(ggrepel)
library(tidyverse)
library(gplots)
library("RColorBrewer")
```

```{r}
#read-count-table

Numers <- '../output/snakemake_out_sf3b1_blood_ds/CountTableNumerators.gz'
Denoms <- '../output/snakemake_out_sf3b1_blood_ds/CountTableDenominators.gz'

NumeratorCountTable <- fread(paste0("gunzip -c ", Numers), sep=" ", header=TRUE, data.table=FALSE)
DenominatorCountTable <- fread(paste0("gunzip -c ", Denoms), sep=" ", header=TRUE, data.table=FALSE)

PSI_Table <- NumeratorCountTable[-1]/DenominatorCountTable[-1]
row.names(PSI_Table) <- gsub(":clu_.+", "", NumeratorCountTable$chrom)

#Replace NaNs from 0/0 operation
PSI_Table[is.na(PSI_Table)] <- 0

#read-leafcutter differntial splicing results
effect.sizes <- read.table(gzfile("../output/snakemake_out_sf3b1_blood_ds/leafcutter.ds.effect_sizes.txt.gz"), header=T)

cluster.significance <- fread(paste0("gunzip -c ", "../output/snakemake_out_sf3b1_blood_ds/leafcutter.ds.cluster_sig.txt.gz"), sep="\t", header=TRUE, data.table=FALSE)

leafcutter.merged <- effect.sizes %>%
  mutate(cluster=gsub("(.+?:).+?:.+?:(clu.+?)", "\\1\\2", intron, perl=T)) %>%
  mutate(junc_id=gsub("(.+?:.+?:.+?):clu.+?_NA", "\\1", intron, perl=T)) %>%
  left_join(cluster.significance, by="cluster")

kable(head(leafcutter.merged))
```


Ok now let's check that the leafcutter differential splicing results make sense. Some diagnositic plots:

```{r}
# how many clusters are there
leafcutter.merged %>%
  distinct(cluster, .keep_all = T) %>% dim()

# and how many are significant (padjust<0.05, 0.01)
leafcutter.merged %>%
  distinct(cluster, .keep_all = T) %>% 
  filter(p.adjust<0.01) %>% dim()

#histogram of pvalues
leafcutter.merged %>%
  distinct(cluster, .keep_all = T) %>% 
  ggplot(aes(x=p)) +
    geom_histogram() +
    theme_bw()
```




```{r}
#volcano plot of cluster-pvalues and largest within-cluster-delta-psi
leafcutter.merged %>%
  mutate(abs.logef = abs(logef)) %>%
  group_by(cluster) %>%
  slice(which.max(abs.logef)) %>%
  mutate(sig=p.adjust<0.05) %>%
  ggplot(aes(x=logef, y=-log10(p), color=sig)) +
    geom_point(alpha=0.05) +
    theme_bw()
```

That doesn't make too much sense, maybe delta-PSI is the better effect size metric to make a sensible volcano plot...


```{r}
# volcano plot of cluster-pvalues and largest within-cluster-delta-psi
leafcutter.merged %>%
  mutate(abs.deltapsi = abs(deltapsi)) %>%
  group_by(cluster) %>% 
  slice(which.max(abs.deltapsi)) %>% 
  mutate(sig=p.adjust<0.05) %>%
  ggplot(aes(x=deltapsi, y=-log10(p), color=sig)) +
    geom_point(alpha=0.05) +
    theme_bw()
```

Now let's check that the junctions identified in Darman et al are enriched for small P-values


```{r}
DarmanJunctionsList <- read.table('../data/other_snakemake_data/DarmanDiffJunctions.LiftedToHg38.bed', header=F)

# Make new id, for easier use in later filtering. Note that coordinates may need to add 1
DarmanJunctionsList$junc_id <- paste(DarmanJunctionsList$V1, DarmanJunctionsList$V2 + 1, DarmanJunctionsList$V3 + 1, sep=":")

NotInDarman<-leafcutter.merged %>% 
  group_by(cluster) %>%
  filter(!any(junc_id %in% DarmanJunctionsList$junc_id)) %>%
  slice(which.min(p))
InDarman<-leafcutter.merged %>% 
  group_by(cluster) %>%
  filter(any(junc_id %in% DarmanJunctionsList$junc_id)) %>%
  slice(which.min(p))

## QQ-plot of cluster P-vals, grouped by whether cluster contains in intron identified as differentially spliced in Daraman et al.
ggplot(NotInDarman, aes(color="NotInDarman", y=-log10(sort(p)), x=-log10(1:length(p)/length(p)))) +
  geom_point() +
  geom_point(data=InDarman, aes(color="InDarman")) +
  xlab("-log10(Theoretical-Pvalues)") +
  ylab("-log10(Observed-Pvalues)") +
  geom_abline() +
  theme_bw() +
  theme(legend.position="bottom") +
  theme(legend.title=element_blank())
```

Ok, now let's check that cryptic upstream 3'ss are enriched

```{r}
#read in annotated 3'ss

```

PCA and heatmap of differentially spliced introns
```{r}
#Filter for max(delta-psi) in top1000 intron clusters for PCA and heatmap
Top1000Clusters

```


