---
title: "20190726_sf3b1_mutWT_k700e_ds"
author: "Stephanie Lozano"
date: "7/26/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

```{r libraries}
library(data.table)
library(dplyr)
library(ggplot2)
library(reshape2)
library(readxl)
library(ggrepel)
library(tidyverse)
library(gplots)
library(knitr)
library("RColorBrewer")
```

To subset leafcutter-generated effect_sizes table for junctions ID'd in Darman et al.

```{r}

#Differential Junctions List - filter for ds junctions listed in Darman et al
#DifferentialJunctionsList <- #read.table('../data/other_snakemake_data/DarmanDiffJunctions.LiftedToHg38.bed', header=F)

# Make new id, for easier use in later filtering. Note that coordinates may need to add 1 -Darman
#DifferentialJunctionsList$junc_id <- paste(DifferentialJunctionsList$V1, #DifferentialJunctionsList$V2 + 1, DifferentialJunctionsList$V3 + 1, sep=":")

```

To subset the leafcutter-generated effect_sizes table for junctions that are identified as significant, based on the leafcutter-generated cluster_sig table.

```{r}
#Differential Junctions List - filter for leafcutter ID'd junctions

#Read in cluster sig text file as data frame and filter for junctions with p value <0.1. 
#SigJcns <- read.delim('../output/snakemake_out_sf3b1_mutWT_k700e_ds/leafcutter.ds.cluster_sig.txt.gz') %>%
  #filter(p.adjust <= 0.05, .preserve= TRUE) 

#Read in effect sizes text file, add chr:clu_ID_- in same format as "cluster" column in SigJcns, filter for chr:cluster_IDs that match those specified in SigJcns data frame.

#DifferentialJunctionsList <- read.delim('../output/snakemake_out_sf3b1_mutWT_k700e_ds/leafcutter.ds.effect_sizes.txt.gz', header=T) %>%
#separate("intron", c("chrom", "chromStart", "chromEnd", "cluster"), ":") %>%
  #mutate(chr_clust= paste(chrom, cluster, sep = ":", collapse = NULL)) %>%
  #filter(chr_clust %in% SigJcns$cluster, .preserve= T)

# Make new id, for easier use in later filtering. Note that coordinates may need to add 1 - leafcutter ID'd jcns
#DifferentialJunctionsList$junc_id <- paste(DifferentialJunctionsList$chrom, #DifferentialJunctionsList$chromStart, DifferentialJunctionsList$chromEnd, sep=":")

```
To subset the leafcutter effect_sizes table for junctions that 1) fall within 3'ss and 2) are significant based on q-values<= 0.05 (i.e hold off on filtering for significant junctions and first filter for junctions that fall within 3'ss)
```{r}

#Differential Junctions List- filter for junctions that fall within 3'ss and are significant based on q-values generated for newly subsetted group


#Read in effect_sizes and reformat as bed file
effect_sizes <- read.delim('../output/snakemake_out_sf3b1_mutWT_e622d_ds/leafcutter.ds.effect_sizes.txt.gz', header=T) %>%
separate("intron", c("chrom", "chromStart", "chromEnd", "cluster"), ":") 

effect_sizes_bed <- effect_sizes %>%
  mutate(chr_clust= paste(chrom, cluster, sep = ":", collapse = NULL)) %>% 
  mutate(strand= str_sub(effect_sizes$cluster, start= -1)) %>%
  mutate(name= ".") %>%
  mutate(score= ".") %>% 
  transform(chromStart = as.numeric(chromStart)) %>%
  transform(chromStart = chromStart - 1) %>%
  transform(chromStart = as.character(chromStart)) %>%
  select(chrom, chromStart, chromEnd, name, score, strand, chr_clust, logef)
  

write.table(effect_sizes_bed,file="../output/snakemake_out_sf3b1_mutWT_e622d_ds/effect_sizes.bed.gz",quote=FALSE, row.names=FALSE,col.names=FALSE,sep="\t")


#Unzip files to be used in python script. Then, run effect_sizes.bed input file through ben's python script and get effect_sizes output file with splice type classifications column

system("gunzip ../data/other_snakemake_data/GencodeHg38_all_introns.bed")
system("gunzip ../output/snakemake_out_sf3b1_mutWT_e622d_ds/effect_sizes.bed")

#Run effect_sizes.bed input file through ben's python script and get effect_sizes output file with splice type classifications column

system("../code/SubSnakemakeWorkflow/scripts/AnnotateSplicingType.py -I ../output/snakemake_out_sf3b1_mutWT_e622d_ds/effect_sizes.bed -A ../data/other_snakemake_data/GencodeHg38_all_introns.bed  -O ../output/snakemake_out_sf3b1_mutWT_e622d_ds/effect_sizes_annotated.bed")

system("gzip ../data/other_snakemake_data/GencodeHg38_all_introns.bed")
system("gzip ../output/snakemake_out_sf3b1_mutWT_e622d_ds/effect_sizes.bed")


#Read in annotated_effect_sizes output, rename columns, add 1 back to chromStart values, and create Alt3ss data fram containing annotated_effect_sizes marked as Alt3ss
annotated_effect_sizes <- read.delim('../output/snakemake_out_sf3b1_mutWT_e622d_ds/effect_sizes_annotated.bed', header=F) 

system("gzip ../output/snakemake_out_sf3b1_mutWT_e622d_ds/effect_sizes_annotated.bed")

names(annotated_effect_sizes) [1] <- "chrom"
names(annotated_effect_sizes) [2] <- "chromStart"
names(annotated_effect_sizes) [3] <- "chromEnd"
names(annotated_effect_sizes) [4] <- "name"
names(annotated_effect_sizes) [5] <- "score"
names(annotated_effect_sizes) [6] <- "strand"
names(annotated_effect_sizes) [7] <- "cluster"
names(annotated_effect_sizes) [8] <- "logef"
names(annotated_effect_sizes) [9] <- "AnnotatedSpliceSite"


Alt3ss <- annotated_effect_sizes %>% filter(AnnotatedSpliceSite == "Alt3ss") %>%
transform(chromStart = as.numeric(chromStart)) %>%
transform(chromStart= chromStart + 1) %>%
transform(chromStart = as.character(chromStart))

#Read in cluster_sig file and subset based on which clusters are annotated as 3'ss in annotated_effect_sizes output

cluster_sig <- read.delim('../output/snakemake_out_sf3b1_mutWT_e622d_ds/leafcutter.ds.cluster_sig.txt.gz', header=T) %>%
  filter(cluster %in% Alt3ss$cluster, .preserve= T)
  

#Generate q-values based on this subset and create SigJcns subset out of samples w/ q-values <= 0.05

  #use p.adjust column from Alt3'ss subset of cluster_sig as input for qvalue fcn. store output of qvalue into qobj data frame

qobj<- qvalue(p = cluster_sig$p.adjust)
cluster_sig$qval= qobj$qvalues

#filter for SigClusters, or clusters w/ qvalue <= 0.05

SigClusters <- cluster_sig %>%
  filter(qval <= 0.05, .preserve= TRUE) 

#DifferentialJunctionsList<- Filter/subset Alt3ss (which is annotated_effect_sizes subset that's annotated as 3'ss) for junctions that match those in SigClusters
  #filter for Alt3ss$cluster that match SigClusters$cluster data frame

DifferentialJunctionsList <- Alt3ss %>% filter(cluster %in% SigClusters$cluster, .preserve = TRUE)

#add junc_id column containing chr:start:stop to later on match to rownames of PSI table.

DifferentialJunctionsList$junc_id <- paste(DifferentialJunctionsList$chrom, DifferentialJunctionsList$chromStart, DifferentialJunctionsList$chromEnd, sep=":")


```
Now we can move on to calculating PSI and generating plots
```{r}
#Calulating PSI
Numers <- '../output/snakemake_out_sf3b1_mutWT_k700e_ds/CountTableNumerators.gz'
Denoms <- '../output/snakemake_out_sf3b1_mutWT_k700e_ds/CountTableDenominators.gz'

NumeratorCountTable <- fread(paste0("gunzip -c ", Numers), sep=" ", header=TRUE, data.table=FALSE)
DenominatorCountTable <- fread(paste0("gunzip -c ", Denoms), sep=" ", header=TRUE, data.table=FALSE)

PSI_Table <- NumeratorCountTable[-1]/DenominatorCountTable[-1]
row.names(PSI_Table) <- gsub(":clu_.+", "", NumeratorCountTable$chrom)

#Replace NaNs from 0/0 operation
PSI_Table[is.na(PSI_Table)] <- 0

# PCA using PSI
PCResults <- (PSI_Table) %>%
  filter(rownames(.) %in% DifferentialJunctionsList$junc_id) %>%
  t() %>%
  prcomp(center=T, scale=T)

PCResults.df <- as.data.frame(PCResults$x[,1:5]) %>% rownames_to_column()

#Read in groups table
Groups <- read.delim("../data/other_snakemake_data/sf3b1_mutWT_e622d_ds.groups.txt", na.strings = "NA", stringsAsFactors = T, header=F)

MutationList <- data.frame(read_excel("../data/other_snakemake_data/SeillerEtAl_MutationList.xlsx"))

# Groups table with rail_ID column

Simplified_Groups<- Groups %>%
  mutate(rail_id= sub(".project2.yangili1.snaptron.processed_to_bedfiles.SRA2.","", V1)) 

#create MDS data frame and use it to add disease_state to Simplified_Groups

MDS<- as.data.frame(c("3641", "18119", "34165",  "28684", "14580"), row.names= NULL) 
colnames(MDS) <- c("rail_id")


Simplified_Groups$disease_state <- ifelse(Simplified_Groups$rail_id %in% MDS$rail_id, "MDS", "CLL")


#Merging Groups (rail_id and V2, or mutation) with PCResults.df into ToPlot

ToPlot <- PCResults.df %>%
  mutate(rail_id = sub(".project2.yangili1.snaptron.processed_to_bedfiles.SRA2.","", rowname)) %>% 
left_join(Simplified_Groups, by= NULL) %>% subset(V1 != "NA")

  
# plot PCResults.df PC1 and PC2, with labels= V2)
ggplot(ToPlot,aes(x=PC1, y=PC2, label=disease_state, color= V2)) +
  # geom_text(alpha=1, size=2.2) +
  geom_point(alpha=0.5) +
  geom_text_repel(size=2.2) +
  theme_bw()

```

It will also be nice to look at a heatmap of this feature space from which the PCA is based on.
```{r}
set.seed(0)
E622D<-ToPlot %>%
  filter(V2 =="K700E") %>%
  dplyr::pull(rowname)

WT<-ToPlot %>%
  filter(V2 =="WT") %>%
  sample_n(7) %>%
  dplyr::pull(rowname)


Heatmap <- PSI_Table %>%
  filter(rownames(.) %in% DifferentialJunctionsList$junc_id) %>%
  dplyr::select(c(E622D, WT)) %>% as.matrix() %>%
  t() %>% scale(scale=F) %>% t() %>% as.data.frame() %>% drop_na() %>% as.matrix()

#
# Heatmap <- t(scale(t(as.matrix(PSI_Table[1:10, c(Cases, Controls)]))))
# Heatmap <- scale(as.matrix(PSI_Table[1:50,c(Cases, Controls)]))


RNAExtractionDate <- as.character(unclass(factor(plyr::mapvalues(colnames(Heatmap), from=ToPlot$rowname, to=ToPlot$V2, warn_missing = FALSE) )))
ColLabel = labCol <- plyr::mapvalues(colnames(Heatmap), from=ToPlot$rowname, to=ToPlot$rail_id, warn_missing = FALSE)

heatmap.2(Heatmap, col=colorRampPalette(c("blue", "white", "red"))(n = 299), dendrogram="column", trace="none", labRow="", labCol=ColLabel, ColSideColors=RNAExtractionDate, key.title="PSI-Zscore", ylab="1593 splice junctions", xlab="10 samples")
```
